Child survival for mothers: mortality change and related measures

### 2020-09-03 paper.tex
not sure

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Child survival for mothers: mortality change and related measures},
  pdfauthor={Ivan Williams and Diego Alburez-Gutierrez},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Child survival for mothers: mortality change and related
measures}
\author{Ivan Williams and Diego Alburez-Gutierrez}
\date{07 August 2019}

@@

Kin count estimation has a long and distinguished pedigree in
mathematical demography, starting with the work of Lotka (1931) on
modelling orphanhood in theoretical populations across demographic
regimes. To the best of our knowledge, Brass (1953) first proposed a
formula to represent child survival over age. The notion of estimating
the expected number of living daughters in a stable population was
generalized by Goodman, Keyfitz, and Pullum (1974) for other kin
relations (granddaughters, cousins, etc.). The ``counting method''
approach was further popularized in Keyfitz and Caswell (2005).
Bongaarts (1987) used a similar approach to estimate descendants in his
`Family Status Model'.

The question of child survival, particularly the survival of daughters
for mothers, sits at the very center of demographic theory. This is
evidenced by the omnipresence of the `net reproductive rate' in accounts
of human and non-human populations. Historical demographers draw
liberally on assumptions about kin availability and individual's
exposure to offspring mortality to explain fertility decisions,
especially around the demographic transition. These assumptions are
often untested given the data sparsity (Livi Bacci 1997; Volk and
Atkinson 2013). Economic theories of fertility that consider children as
`investment' for old-age support also rely on notions of child survival
(Preston and others 1978).

Child survival - and it's counterpart, child death - gain new
significance in the wake of the fertility transition. Depressed
fertility means that fewer children are tasked with providing key
emotional, social, and financial transfers to aging parents for ever
increasing periods of time. With increasing periods of generational
overlap, individuals find themselves `sandwiched' between aging parents
and young children requiring their simultaneous attention and care
(Daatland, Veenstra, and Lima 2010). In the context of global population
aging, elderly parents without access to formal social security and
pension systems are particularly reliant on these transfers to make ends
meet (Smith-Greenaway and Trinitapoli 2020).\\
Losing an only child may become a more common experience in the context
of global fertility decline, something particularly worrying for parents
suddenly rendered childless. Given the known psychological, health, and
social consequences of child loss for bereaved parents and families
(Hendrickson 2009; Lee et al. 2014), it is surprising that the
demographic processes that shape parental bereavement remain very poorly
understood. Recent development have underscored the need to understand
how sudden changes in mortality affect the availability of kin. In the
context of the global pandemic caused by the Covid-19 disease, elderly
people depend on support provided by their relatives, especially during
periods of lockdown and in the absence of governmental support
mechanisms. Elderly parents may be at a higher risk of losing the key
support provided by their adult children given the known age-gradient in
the Covid-19 case fatality rate. How do changes in mortality affect the
availability of offspring over age from the point of view of a
prospective mother? How can we characterize the timing of offspring
mortality as experienced by a mother from a formal demographic
perspective? How this would work in an heterogeneity population? In the
remainder of the paper, we aim to formalize the relationship
population-level changes in mortality rates and perceived change in the
lived experience of child death from the perspective of a mother. Our
approach is key for understanding the impact of age-specific excess
mortality on the resilience or otherwise of kinship networks for the
vulnerable elderly population. We consider the case when these changes
are additive, mutliplicative, and when they are not evenly distributed
across age. In this paper, we uncover formal relationships to answer
these questions and provide numerical applications using data from
countries in Latin America during the second half of the twentieth
century. See the \protect\hyperlink{Data}{data section} in the
\protect\hyperlink{Applications}{Applications} section for more details
on the source, calculation procedures and main results. We used period
measures, in an stable context, so the objetive is give moment measures.
A more sophisticated model using the appropriate age-specific cohort
rates in the subscripts has been proposed by Alburez-Gutierrez, Kolk,
and Zagheni (2019).

\hypertarget{relationship-proof}{%
\subsection{Relationship \& Proof}\label{relationship-proof}}

@@

\[
\begin{aligned}
CS_a = \int_{0}^a{m_{x}l_{a-x}dx}
\end{aligned}
\]

@@

As an initial step, we seek an intuitive understanding of the effect of
changing mortality on child survival. Building on work by Keyfitz and
Caswell (2005) for the probability of a living mother, we show which
features of the child survival function explain the expected number of
children. For this we use an approximate of \(l_x\) using Taylor´s
theorem until second order around the mean age of childbearing
\(\kappa\) for all \(a>=\alpha\):

\[CS_a \approx l_{a-\kappa} \int_{\alpha}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]\]

where the fertility pattern by age is concentrated around \(\kappa\) and
the accumulated fertility (or gross reproduction rate in our same-sex
model) is \(F_a = \int_{\alpha}^a{m_{x}dx}\). The second Taylor´s term
is null because \(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\).

We find that, seen from the perspective of a mother, child survival
mainly depends on the cummulative fertility function\footnote{Note that
  this runs contrary to the findings of Keyfitz and Caswell (2005), who
  found that levels of fertility did not affect the probability of
  having living ancestors.} and the mean age at childbearing \(\kappa\)
acts through the factor \(l_{a-\kappa}\). The dispersion of of fertility
\(\sigma^2\) is negatively proportional to child survival. All else
constant, a higher dispersion of fertility \(\sigma^2\) is associated
with an increase in \(CS_a\) during reproductive age. The negative
relationship stems from the fact that the second derivative of
\(l_{a-\kappa}\) is negative in the 15-49 age range, given a concave
downwards survival function.

The second Taylor´s term is null because
\(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\). The formula tells us
that the survival experience for a woman aged \emph{a} depends mainly of
cumulated fertility (possible cases, different from the ancestor count
case where this doesn´t affect) and child survival probability (success
cases). Also that the approximation is affected by how disperse is the
fertility by age (variance \(\sigma^2\)) and the suvival curvature
around (with negative sign), in a \(l_x\) range where is tipycally very
flat (range 20-40 years old) in transitioned populations.

And change in \(\kappa\), taking logs and deriving:
\(\frac{dCS_a}{d\kappa}\frac{1}{CS_a} \approx \mu_{a-\kappa}F_a \Delta\kappa\)

@@

We now consider the consequences of an absolute change \(\delta\) in
mortality in the range \([0,a-\alpha]\), where \(\alpha\) is the start
age of fertility risk. As a \protect\hyperlink{Corollary}{corollary} we
consider the case where mortality only changes for infants, considering
that it is unlikelily for a potential change in mortality to affect all
ages equally in real populations. For now, let us consider
\(m_{x,\delta}=m_{x}+\delta\) (Wrycza and Baudisch (2012)) and
\(l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}\):

@@

\[\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}\]

Dividing both sides by \(CS_{a}\) in an discrete approximation we get

@@

The expected change in descendants survival is inversely proportional to
the difference between maternal age \(a\) and the mean age of the mother
at the birth of her surviving daughters \(k_a\) (always smaller than
\(\kappa_a\)). The magnitude of the change depends negatively on the age
distribution of the surviving offspring. This is intuitive considering
that younger descendants experiences longer periods of exposure to risk.
The figure @ref(fig:CS\_abs\_app) shows that goodness of fit is
decreasing with the change size, given that \(\delta\) is assumed near
zero (as an example considering \(\delta=0.01\) means at age 50 that
\(\frac{l_50}{e^{\delta\,50}} = \frac{l_50}{1.6}\), a big change).

That an equal change in mortality could happens at all ages is
unlikelihood. This measure is much more related to first ages (child and
youth) than adult or ones. As a corollary, if the change only happens in
the age range \(\left[0;1\right)\), infant mortality, we can inspect the
effect starting with splitting the integral (of course, for \(a>1\)){]}:

\[CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx\]

Deriving by \(\delta\) and valuating near 0, we get:

\[CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}\]
If we name \(CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}\), we can
express the first term in the right as \(CS_a - CS_{a-1,a}\), and
finally divide by \(CS_a\):
\[ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta\]

This means that an absolute change in infant hazard rate affects
proportionally for all the age range extension (because of all the
cohorts aged more than 1), except for the mean time exposure for those
who born and survive in the last year \(1-(a-k_{a-1,a})\), times
\(\frac{CS_{a-1,a}}{CS_a}\) the portion of cummulative alive daughters
for the same period, weighting how important was this in total
successful experience (in terms of alive descendants).

@@

\[\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt\]

The last integral is \(CS_{a-t}\). Dividing by \(CS_a\) and multiplying
by \(\delta\) get one useful expression. This is the negative of
cumulative hazard \(H_a\) but considering a positive factor
\(\frac{CS_{a-t}}{CS_a}\) that takes in account the relative amount of
surviving descendants that would be lost at each child age at risk.

@@

Grasping this relationship intuitively is more difficult given the
interaction of birth and mortality rates (Keyfitz and Caswell (2005)).
This factor has a S-shape because \(l_x\) curvature and fertility
accumulation, that gives more weight to first ages, generally less than
1 except for mother ages close to age valuation \emph{a} with complete
reproductive life and high infant mortality (no additional children and
high risk for one age additional of survive), typically some countries
with \(a=50\) and first periods.\\

@@

We call \emph{Mean Time Spent in Bereavement} (\emph{MTSB}) the absolute
measure of the expected total lifetime with a death daughter for a
mother aged \emph{a}, which can be expressed in terms of a temporary
expected lost years index in line with \(e^\dagger\) (Vaupel (1986)):

@@

Where \(d_t\) is the death distribution from birth, \(e_{0|a-x-t}\) is
the life expectancy at birth until age \(a-x-t\) and
\(e_{0|a-x}^\dagger\) the temporary dispersion measure. But most
interesting would be to compare these years with the time that mothers
would expect to live with their daughters. We call this the
\emph{Intensity Time in Bereavement} (\(ITB\))): a ratio between
expected time with a ``lost'' and expected time with a ``life'', that
allows to make comparisons between population regimes.

@@

This is a ratio between child-years in two radically different states.
Looks similar to the transcendental entropy measure \(H\) (Keyfitz and
Caswell (2005)) but considering all the cohorts born during the the
mother´s life, weighted by their relative size \(m_x\). In the figure
@ref(fig:plot\_ITL\_MAL) is shown that \emph{ITB} is bigger for young
women because of the weight of infant mortality in their mother
experience, also with more dispersion for same levels of parity.

Another important factor for the child survival experience of mothers,
is the mean age at child loss, called here \(MAL\). This relation can be
derived by starting with the mother age \(x+t\) at each death child age
\(t\) at death, weighted by the fertility and survival function. In it,
\(\kappa\) is the mean age at childbirth for women aged \(a\),
\(MAD_{a-x}\) refers to the mean age at death for newborns that die
before \(a-x\), \(F_a\) is the accumulated fertility for a women aged
\(a\):

\[MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}\]

\hypertarget{heterogeneity}{%
\subsubsection{Heterogeneity}\label{heterogeneity}}

In a model with heterogeneity, determined at birth with a multiplicative
effect, \(CS_a\) could be interpreted as a conditional expectation with
random variables \emph{K} for fertility heterogeneity and \emph{Z} for
mortality frailty.

@@

Following Coresh and Goldman (1988) we can express the fertility part in
a multiplicative way with variability in the level but not in the shape,
as \({m(k)}_x = F_\beta\,k\,r_x\), where \(F_\beta\) is the baseline
cumulated fertility until upper limit in fecundity age range \(\beta\),
\(K\) is a random variable with mean 1 that allows variability between
groups and \(r_x\) is the fertility structure by age
(\(\int_{0}^{\beta}{r_x}=1\)). Frailty part can be thought in a cohort
effect way, also in the multiplicative assumption as Vaupel and Missov
(2014), with \(l(z)_x=e^{-H_{x}Z}=l_x^Z\), with baseline hazard \(\mu\).
Considering the joint distribution \(f_{kz}\), the intuition could be
expressed as \(f_{zk}=f_{k|z}f_z\): groups with higher descendant´s
mortality would adjust their fertility level, with positive correlation.
Replacing both variables, the unconditional mean \(\overline{CS}_a\)
would be\footnote{Note that given K and Z values fertility and child
  survival are independent.}:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Child survival for mothers: mortality change and related measures},
  pdfauthor={Iván Williams and Diego Alburez-Gutierrez},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Child survival for mothers: mortality change and related measures}
\author{Iván Williams and Diego Alburez-Gutierrez}
\date{May 2020}

@@

The question of child survival for mothers sits at the very center of
demographic theory. Offspring survival is usually studied alongside
fertility decline in the context of the demographic transition, but it
also matters for post-transitional societies. Lower fertility and higer
life expectancy means that fewer children are expected to provide key
emotional, social, and financial transfers to aging parents for longer
periods of time. With increasing periods of generational overlap,
individuals find themselves `sandwiched' between aging parents and young
children requiring their simultaneous attention and care (Daatland,
Veenstra, and Lima 2010). In the context of global population aging,
elderly parents without access to formal social security and pension
systems are particularly reliant on these transfers (Smith-Greenaway and
Trinitapoli 2020). Given the known psychological, health, and social
consequences of child loss for bereaved parents and families
(Hendrickson 2009; Lee et al. 2014), it is surprising that the
demographic processes that shape parental bereavement remain very poorly
understood.

The Covid-19 pandemic has underscored the need to understand how sudden
changes in mortality affect the availability of kin. The elderly depend
on kin support, especially during periods of lockdown and in the absence
of governmental support mechanisms. Elderly mothers may be at a higher
risk of losing the key support provided by their adult children given
the known age-gradient in the Covid-19 case fatality rate. How do
changes in mortality affect the availability of offspring over age from
the point of view of a prospective mother? How do these processes play
out in a heterogeneous population?

In this paper, we aim to formalize the relationship between
population-level changes in mortality rates (additive or multiplicative)
and changes in the lived experience of kin death. This is key for
understanding the impact of age-specific excess mortality on the
resilience or otherwise of kinship networks for the vulnerable elderly
population. This paper uncovers formal relationships and provides
numerical applications using data from Latin America in the second half
of the twentieth century using period measures in a stable
context.\footnote{See the \protect\hyperlink{Applications}{Applications}
  section for more details on the source, calculation procedures and
  main numerical results.} We continue a long tradition in mathematical
demography, starting with the work of Lotka (1931) on modelling
orphanhood in theoretical populations, Brass (1953) on child survival,
Goodman, Keyfitz, and Pullum (1974) on kin survival in stable
populations, and recently, Caswell (2019) kinship matrices.

\hypertarget{relationship-and-proofs}{%
\subsection{Relationship (and proofs)}\label{relationship-and-proofs}}

@@

\[CS_a = \int_{0}^a{m_{x}l_{a-x}dx}\]

@@

As an initial step, we seek an intuitive understanding of the previous
relation. Building on work by Keyfitz and Caswell (2005) for the
probability of a living mother, we can show which features of the child
survival function explain the expected number of children. For this we
use an approximate of \(l_x\) using Taylor´s theorem until second order
around the mean age of childbearing \(\kappa\):

\[
\begin{aligned}
CS_a \approx & l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  & \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  & \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]
\end{aligned}
\]

where the fertility pattern by age is concentrated around \(\kappa\) and
the accumulated fertility (or gross reproduction rate in our
female-dominant scenario) is \(F_a = \int_{\alpha}^a{m_{x}dx}\). The
second Taylor´s term is null because
\(\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a\).

We find that, seen from the perspective of a mother, child survival
mainly depends on the cummulative fertility function and the survival of
daughters from birth to \(a-\kappa\). The approximation is affected
negativally by how disperse is the fertility by age (variance
\(\sigma^2\)) and the survival curvature around (with negative sign), in
an age range where \(l_x\) is tipycally very flat in transitioned
populations (range 20-40 years old)\footnote{This approximation is
  useful also for get an idea of impact because of changes in the
  fertility average age \(\kappa\), keeping constant the level: taking
  logs in \(F_a \, l_{a-\kappa}\) and deriving we get
  \(\frac{\Delta CS_a}{CS_a} \approx \mu_{a-\kappa} \Delta\kappa\),
  being daughters exposed to less time beacuse of the delay in pattern}.

@@

We now consider the consequences of an absolute change \(\delta\) in
mortality in the range \([0,a-\alpha]\), where \(\alpha\) is the start
age of fertility risk. For now, let us consider
\(m_{x,\delta}=m_{x}+\delta\) (Wrycza and Baudisch (2012)) and
\(l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}\):

@@

\[
\begin{aligned}
\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}
\end{aligned}
\]

Dividing both sides by \(CS_{a}\) in an discrete approximation we get:

@@

The expected change in descendants survival is inversely proportional to
the difference between maternal age \(a\) and the mean age of the mother
at the birth of her surviving daughters \(k_a\). The magnitude of the
change depends negatively on the age distribution of the surviving
offspring (the younger the fertility bigger the impact), which is
intuitive considering that older descendants experiences longer periods
of exposure to risk.

However, it is unlikely for mortality to change at the same rate at all
ages. In the course of the demographic transition, for example, we would
expect larger changes in young-age mortality. Consider a change in
infant mortality in the age range \(\left[0;1\right)\). We can inspect
the effect in child survival \(CS_{a}^{\delta_0}\) by splitting the
integral between those daughters who were exposed to the mortality
change and those who were not (now generalizing for \(\alpha\) positive
and \(a>1\) for simplicity):

\[
\begin{aligned}
CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx \\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx
\end{aligned}
\] Deriving by \(\delta\) and valuating near 0, we get:

\[\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}\]
Given that \(CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}\), we can
express the first term in the right as \(CS_a - CS_{a-1,a}\). The
righ-most term could also be expressed as \(\kappa_{a-1,a} CS_{a-1,a}\),
so that:

\[ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\]
The factor \(1-(a-\kappa_{a-1,a})\), which lies between 0 and 1, is
equal to the average portion of time spent with living daughters durings
age \emph{a}. Assuming a uniform distribution of those births,
discretizing the change and dividing by \(CS_a\) we get the final
expression for change in infant mortality:

\[\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a}\frac{1}{2}\right] \Delta\delta\]

This means that an absolute change in infant hazard rate affects
proportionally for all the age range extension (cohorts aged more than
1), except for ``half'' the portion \(\frac{CS_{a-1,a}}{CS_a}\) on how
important was this las age in total ``successful'' experience (in terms
of alive descendants). If \(a\) is near \(\beta\) then
\(\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta\).

@@

\[
\begin{aligned}
\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt
\end{aligned}
\]

Considering that the last integral is equal to \(CS_{a-t}\), we divide
by \(CS_a\) and multiply by \(\delta\) to obtain:

@@

This is the negative of cumulative hazard \(H_a\) but considering a
positive factor \(\frac{CS_{a-t}}{CS_a}\) that takes in account the
relative amount of surviving descendants that would be lost at each
child age at risk, that gives more weight to first ages because all the
parity was exposed in that age.

Grasping this relationship intuitively is more difficult given the
interaction of birth and mortality rates (Keyfitz and Caswell (2005)).
The factor \(\frac{CS_{a-t}}{CS_a}\) has a S-shape because \(l_x\)
curvature and fertility accumulation. Taking the difference between
consecutive ages \(a\) and \(a+1\), we have
\(CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}\),
\(CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}\). In the last
one, a second term is added because of additional born daughters after
\(a-1\), and in the first term the existing one are survived at on
\(\frac{l_{a-x}}{l_{a-1-x}}\). In advanced ages when fertility is
complete, the difference is only explained by the surviving part, that
is why this factor is not a strictly increasing function but for mother
ages close to \(\beta\) (last theoretical fecundity age). We provide a
numerical evaluation of both analytical expressions (absolute an
relative) in the \protect\hyperlink{Applications}{Applications} section.

@@

We call \emph{Mean Time Spent in Bereavement} (\emph{MTSB}) the absolute
measure of the expected total lifetime with a death daughter for a
mother aged \emph{a}, which can be expressed in terms of a temporary
expected lost years index, in line with \(e^\dagger\) (Vaupel (1986)):

@@

Where \(d_t\) is the death distribution from birth, \(e_{0|a-x-t}\) is
the life expectancy at birth until age \(a-x-t\) and
\(e_{0|a-x}^\dagger\) the temporary dispersion measure. But most
interesting would be to compare these years with the time that mothers
would expect to live with their daughters. We call this the
\emph{Intensity Time in Bereavement} (\(ITB\)): a ratio between expected
time with a ``lost'' and expected time with a ``life'', that allows to
make comparisons between population regimes.

@@

This is a ratio between child-years in two radically different states.
Looks similar to the transcendental entropy measure \(H\) (Keyfitz and
Caswell (2005)) but considering all the cohorts born during the mother´s
life, weighted by their relative size \(m_x\). In the figure
@ref(fig:plot\_ITL\_MAL) is shown that \emph{ITB} is bigger for young
women because of the weight of infant mortality in their mother
experience, also with more dispersion for same levels of parity.

Another important factor for the child survival experience of mothers,
is the mean age at child loss, called here \(MAL\). This relation can be
derived by starting with the mother age \(x+t\) at each death child age
\(t\) at death, weighted by the fertility and survival function. In it,
\(\kappa\) is the mean age at childbirth for women aged \(a\),
\(MAD_{a-x}\) refers to the mean age at death for newborns that die
before \(a-x\) and \(F_a\) is the accumulated fertility for a women aged
\(a\):

\[
\begin{aligned}
MAL_a = & \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = & \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = & \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = & \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}
\end{aligned}
\]

Following this, in populations with high infant mortality
\(MAD_{a-x}<1\) for all \emph{x}, so \(MAL_a\approx\kappa_a+f_0\), being
\(f_0\) the average time spent for those child that dies in their first
year of life. A numerical approximation of both measures are done in
\protect\hyperlink{Applications}{Applications} section.

\hypertarget{heterogeneity-first-ideas}{%
\subsubsection{Heterogeneity (first
ideas)}\label{heterogeneity-first-ideas}}

In a model with heterogeneity, determined at birth by a multiplicative
effect, \(CS_a\) could be interpreted as a conditional expectation with
random variables \emph{K} for fertility heterogeneity and \emph{Z} for
mortality frailty.

@@

Following Coresh and Goldman (1988) we can express the fertility part in
a multiplicative way with variability in the level but not in the shape,
as \({m(k)}_x = F_\beta\,k\,r_x\), where \(F_\beta\) is the baseline
cumulated fertility until upper limit in all age range until \(\beta\),
\(k\) is a random variable with mean 1 that allows variability between
groups and \(r_x\) is the fertility structure by age
(\(\int_{0}^{\beta}{r_x}=1\)). In the other side, frailty part can be
thought in a cohort effect way, also in the multiplicative assumption as
Vaupel and Missov (2014), with \(l(z)_x=e^{-H_{x}z}=l_x^z\), with
baseline hazard \(\mu\). Considering the joint distribution \(f_{kz}\),
the intuition could be expressed as \(f_{zk}=f_{k|z}f_z\): groups with
higher descendant´s mortality would adjust their fertility level, with
positive correlation. Replacing both variables, the unconditional mean
\(\overline{CS}_a\) would be\footnote{Note that given k and z values,
  fertility and child survival are independent.}:

\[
\begin{aligned}
\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx
\end{aligned}
\]
